---
title: "Census Data Cleaning"
output: pdf_document
date: "2025-01-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

```{r}
library(dplyr)
library(stringr)
library(janitor)
```
```{r}
race.raw <- read.csv("Census 2020 Race by Tract.csv")
age <- read.csv("Census ACS 2023 5-year est Age and Sex by Tract.csv")
poverty <- read.csv("Census ACS 2023 5-year est Poverty by Tract.csv")
vacancy <- read.csv("Census 2020 Vacancy by Block.csv")
```

```{r}
# Transpose race data
race <- t(race.raw)
race <- as.data.frame(race)
# Convert row names to tract column
race <- data.frame(NAME = rownames(race), race, row.names = NULL)
```

```{r}
# Standardize tract names

extract_tract_number <- function(name) {
  # Extract the "Census Tract \\d+\\.?\\d*" portion
  tract_portion <- str_extract(name, "Census.?Tract.?\\d+\\.?\\d*")
  
  # If found, extract just the numeric tract number
  if (!is.na(tract_portion)) {
    tract_number <- str_extract(tract_portion, "\\d+\\.?\\d*")
    
    # Remove any leading zeros
    tract_number <- str_replace_all(tract_number, "^0+", "")
    # Command to remove trailing period from race tracts???
    
    return(tract_number)
  }
  
  # If not found, return NA
  return(NA_character_)
}

extract_census_tract <- function(df) {
  df$tract_number <- sapply(df$NAME, extract_tract_number)
  df <- dplyr::relocate(df, tract_number, .before = 1)
  df$tract_number <- gsub("\\.", "", df$tract_number)
  return(df)
}

race <- extract_census_tract(race)
age <- extract_census_tract(age)
poverty <- extract_census_tract(poverty)
vacancy <- extract_census_tract(vacancy)

```

```{r}
# Rename vacancy columns and remove extraneous columns

vacancy <- vacancy[, -which(names(vacancy) == "GEO_ID")]
vacancy <- vacancy[, -which(names(vacancy) == "NAME")]
vacancy <- vacancy[, -which(names(vacancy) == "X")]

vacancy <- vacancy[-c(1), ]
colnames(vacancy) <- c("tract_number", "Total","Occupied","Vacant")

# Summarize vacancy data based on tract

vacancy <- vacancy %>%
  mutate(
    Total = as.numeric(Total),
    Occupied = as.numeric(Occupied),
    Vacant = as.numeric(Vacant)
  )
vacancy <- vacancy %>%
  group_by(tract_number) %>%
  summarize(
    Total = sum(Total, na.rm = TRUE),
    Occupied = sum(Occupied, na.rm = TRUE),
    Vacant = sum(Vacant, na.rm = TRUE)
  )
```

```{r}
# Rename race columns and remove extraneous columns
colnames(race) <- race[1,]
race <- race[-1, ] 
colnames(race)[1] <- "tract_number"
# Rename age columns and remove extraneous columns
colnames(age) <- age[1,]
age <- age[-1, ] 
colnames(age)[1] <- "tract_number"
# Rename poverty columns and remove extraneous columns
colnames(poverty) <- poverty[1,]
poverty <- poverty[-1, ] 
colnames(poverty)[1] <- "tract_number"
```

```{r}
# Horizontally combine dataframes based on tract

combined_data <- race %>%
  full_join(vacancy, by = "tract_number") %>%
  full_join(poverty, by = "tract_number") %>%
  full_join(age, by = "tract_number")
```

```{r}
# Drop columns with lots of missing values 

missing_data <- colSums(is.na(combined_data)) / nrow(combined_data) * 100

# Define thresholds
high_missing_threshold <- 50  # Columns with more than 50% missing values
low_variance_threshold <- 1   # Columns with only one unique value

# Identify columns to drop
columns_to_drop <- names(combined_data)[missing_data > high_missing_threshold]

# Drop unnecessary columns
combined_data <- combined_data %>%
  select(-all_of(columns_to_drop))

# Drop margin of error columns
combined_data <- combined_data %>%
  select(-contains("Margin of Error"))
```

```{r}
# Find latitude and longitude to correspond to each census tract
library(tidyr)

# .txt from Census (https://www.census.gov/geographies/reference-files/time-series/geo/centers-population.html)converted to .csv
tractcoords <- read.csv("CenPop2020_Mean_TR42.csv")

# Separate single column into six columns
tractcoords <- tractcoords %>%
separate(
    col = 1,  # First column
    into = c("STATEFP", "COUNTYFP", "TRACTCE", "POPULATION", "LATITUDE", "LONGITUDE"),
    sep = ",",
    convert = TRUE
  ) %>%
  mutate(
    LATITUDE = as.numeric(str_remove(LATITUDE, "\\+")),
    LONGITUDE = as.numeric(LONGITUDE)
  )

# Rename tract name column
colnames(tractcoords)[colnames(tractcoords) == "TRACTCE"] <- "tract_number"

tractcoords$tract_number <- as.numeric(tractcoords$tract_number)
combined_data$tract_number <- as.numeric(combined_data$tract_number)

combined_data <- combined_data %>%
  full_join(tractcoords, by = "tract_number")

```

```{r}
# Save combined dataset
write.csv(combined_data, "combined_data.csv", row.names = FALSE)
cleaned_data <- read.csv("cleaned_data.csv")
```

# Select relevant columns
#cleaned_data <- subset(combined_data, select = c(
  "tract_number", "Label..Grouping.", "Total",
  "Hispanic or Latino", "Not Hispanic or Latino",
  "White alone", "Black or African American alone",
  "American Indian and Alaska Native alone", "Asian alone",
  "Native Hawaiian and Other Pacific Islander alone",
  "Some Other Race alone", "Population of two races",
  "Total", "Vacant", "Occupied",
  "Estimate!!Total!!Population for whom poverty status is determined",
  "Estimate!!Total!!Total population!!AGE!!Under 5 years",
  "Estimate!!Total!!Total population!!AGE!!5 to 9 years",
  "Estimate!!Total!!Total population!!AGE!!10 to 14 years",
  "Estimate!!Total!!Total population!!AGE!!15 to 19 years",
  "Estimate!!Total!!Total population!!AGE!!20 to 24 years",
  "Estimate!!Total!!Total population!!AGE!!25 to 29 years",
  "Estimate!!Total!!Total population!!AGE!!30 to 34 years",
  "Estimate!!Total!!Total population!!AGE!!35 to 39 years",
  "Estimate!!Total!!Total population!!AGE!!40 to 44 years",
  "Estimate!!Total!!Total population!!AGE!!45 to 49 years",
  "Estimate!!Total!!Total population!!AGE!!50 to 54 years",
  "Estimate!!Total!!Total population!!AGE!!55 to 59 years",
  "Estimate!!Total!!Total population!!AGE!!60 to 64 years",
  "Estimate!!Total!!Total population!!AGE!!65 to 69 years",
  "Estimate!!Total!!Total population!!AGE!!70 to 74 years",
  "Estimate!!Total!!Total population!!AGE!!75 to 79 years",
  "Estimate!!Total!!Total population!!AGE!!80 to 84 years",
  "Estimate!!Total!!Total population!!AGE!!85 years and over"
))

